---
title: "Implementing Gaussian Process in R and Python"
author: Hao Zhu
date: "2019-09-12"
output: html_document
slug: gp_python_r
categories:
  - Deep Learning
tags:
  - R
  - Python
  - Deep Learning
  - Gaussian Process
---

```{r}
library(reticulate)
reticulate::use_condaenv("dl")
knitr::opts_chunk$set(message = F, warning = F)
```

This blog post is trying to implementing Gaussian Process (GP) in both Python and R. The main purpose is for my personal practice and hopefully it can also be a reference for future me and other people. In fact, it's actually converted from my first homework in a [Bayesian Deep Learning class](https://www.cs.tufts.edu/comp/150BDL/2019f/). 

A interesting finding is that the implementation in R is much faster than the one using `numpy`. 

All of the equations or figures mentioned in this post can be referened in the [Rasmussen & Williams' textbook for Gaussian Process](http://www.gaussianprocess.org/gpml/chapters/).


# Background
Gaussian Process (GP) can be represented in the form of

$f(\mathbf{x}) \sim \mathcal{GP}(m(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$

where $m(\mathbf{x})$ is the mean function and $k(\mathbf{x}, \mathbf{x'})$ is the covariance/kernel function. In this post, we are trying to create some kernel functions from scratch. We will also create methods to sample values from the prior and the posterior. 

# Mean & kernel functions
For simplicity, our mean function is set to be 0 for all x inputs.

$m(\textbf{x}) = 0$

Squared Exponential kernel (Eq. 4.9)

$k_{SE}(r) = exp(-\frac{r^2}{2l^2})$

Matern kernel (Eq. 4.14)

$k_{Matern}(r) = \frac{2^{1-\nu}}{\Gamma(\nu)}(\frac{\sqrt{2\nu}r}{\ell})^{\nu}K_{\nu}(\frac{\sqrt{2\nu}r}{\ell})$

<div style="display: inline-block;vertical-align:top; width: 48%;">

## Implementations in Python
```{python}
import numpy as np
from scipy.special import gamma, kv
import matplotlib.pyplot as plt

def mean_func(x):
    return np.zeros(len(x))
    
def get_r(x1, x2):
    return np.subtract.outer(x1, x2)

def sqexp_kernel(r, l = 1):
    return np.exp(-0.5 * (r/l)**2)

def matern_kernel(r, l = 1, v = 1):
    r = np.abs(r)
    r[r == 0] = 1e-8
    part1 = 2 ** (1 - v) / gamma(v)
    part2 = (np.sqrt(2 * v) * r / l) ** v
    part3 = kv(v, np.sqrt(2 * v) * r / l)
    return part1 * part2 * part3
```

</div>
<div style="display: inline-block;vertical-align:top; width: 48%;"> 

## Implementations in R
```{r, message=F, fig.height=3}
library(MASS)
library(plotly)

mean_func <- function(x) {
  rep(0, length(x))
}

get_r <- function(x1, x2) {
  outer(x1, x2, FUN = "-")
}

sqexp_kernel <- function(r, l = 1) {
  exp(-0.5 * (r/l)^2)
}

matern_kernel <- function(r, l = 1, v = 1) {
  r <- abs(r)
  r[r == 0] <- 1e-8
  part1 = 2^(1-v) / gamma(v)
  part2 = (sqrt(2*v) * r / l)^v
  part3 = besselK(sqrt(2*v) * r / l, nu = v)
  return(part1 * part2 * part3)
}
```

</div>

Now let's try to recreate the input-distance to covariance figure  using the functions we defined here. Our result (done in python for my homework) is the same as the figures (e.g. Figure 4.1) in the R&W textbook. 

```{python, echo = F}
sqexp_x1 = np.zeros(30)
sqexp_x2 = np.linspace(0, 3, 30)
sqexp_r = sqexp_x2 - sqexp_x1

plt.figure(figsize = (10, 3))
plt.subplot(121)
plt.plot(sqexp_x2, sqexp_kernel(sqexp_r, 0.5), 'b', label = "$\\ell$=0.5")
plt.plot(sqexp_x2, sqexp_kernel(sqexp_r, 2), 'r:', label = "$\\ell$=2")
plt.plot(sqexp_x2, sqexp_kernel(sqexp_r, 10), 'g:', label = "$\\ell$=10")
plt.legend(loc = "upper right")
plt.xlabel("input distance, r")
plt.ylabel("covariance, k(r)")
plt.title("Squared Exponential kernel")

matern_x1 = np.zeros(30)
matern_x2 = np.linspace(0.00001, 3, 30)
matern_r = matern_x2 - matern_x1

plt.subplot(122)
plt.plot(matern_x2, matern_kernel(matern_r, 1, 0.5), 'b', label = "$\\ell$=1,$\\nu$=0.5")
plt.plot(matern_x2, matern_kernel(matern_r, 1, 2), 'r:', label = "$\\ell$=1,$\\nu$=2")
plt.plot(matern_x2, matern_kernel(matern_r, 1, 10), 'g:', label = "$\\ell$=1,$\\nu$=10")
plt.legend(loc = "upper right")
plt.xlabel("input distance, r")
plt.ylabel("covariance, k(r)")
plt.title("Matern kernel")
plt.show()
```

## Comparing the kernel performance

<div style="display: inline-block;vertical-align:top; width: 48%;">
### Python performance
```{python}
from time import process_time
x1 = np.random.rand(100)
x2 = np.random.rand(100)
start = process_time()
for i in range(50): sqexp_test = sqexp_kernel(get_r(x1, x2))
print("SE kernel for rand:" + str((process_time() - start) / 50), " s")
start = process_time()
for i in range(50): matern_test = matern_kernel(get_r(x1, x2))
print("Matern kernel for rand:" + str((process_time() - start) / 50), " s")
```

</div>

<div style="display: inline-block;vertical-align:top; width: 48%;">
### R performance
```{r}
x1 = runif(100)
x2 = runif(100)
start = Sys.time()
for(i in 1:50) se_test = sqexp_kernel(get_r(x1, x2))
print(paste("SE kernel for rand:", (Sys.time() - start)/ 100, "s"))
start = Sys.time()
for(i in 1:50) matern_test = matern_kernel(get_r(x1, x2))
print(paste("Matern kernel for rand:", (Sys.time() - start)/ 100, "s"))
```

</div>

The results are quite comparable. I also checked the performance when it scales up, it's still quite similar. In fact, especially for Matern kernel, when the size of the input vectors get big, I feel like it's slightly faster to do it in R.

# Sampling function
## Sampling from the Prior

<div style="display: inline-block;vertical-align:top; width: 48%;">

### Implementations in Python
```{python}
def sample_prior(
        x, mean_func, cov_func, cov_args = {}, 
        random_seed = -1, n_samples = 5):
    x_mean = mean_func(x)
    x_cov = cov_func(get_r(x, x), **cov_args)
    random_seed = int(random_seed)
    if random_seed < 0:
        prng = np.random
    else:
        prng = np.random.RandomState(random_seed)
    out = prng.multivariate_normal(x_mean, x_cov, n_samples)
    return out
```

</div>

<div style="display: inline-block;vertical-align:top; width: 48%;">

### Implementations in R
```{r}
sample_prior <- function(
  x, mean_func, cov_func, ..., random_seed = -1, n_samples = 5
) {
  x_mean <- mean_func(x)
  x_cov <- cov_func(get_r(x, x), ...)
  random_seed <- as.integer(random_seed)
  if (random_seed > 0) set.seed(random_seed)
  return(MASS::mvrnorm(n_sample, x_mean, x_cov))
}
```

</div>

Let's try to get a few samples from the prior with SE kernel at different length-scales $\ell$. This time, let's do it in python. 

```{python}
plt.figure(figsize = (10, 3))
x = np.linspace(-20, 20, 200)
for l_i, l in enumerate([0.25, 1.0, 4.0]):
    y = sample_prior(x, mean_func, sqexp_kernel, 
                     {"l": l}, n_samples = 5)
    plt.subplot(131 + l_i)
    for i in range(5):
        plt.plot(x, y[i, :], "-", alpha = 0.3)
    plt.title("$\\ell$ = " + str(l))
```

## Sampling from the Posterior
Here we will implement "Prediction using Noisy Observations" because the Noise-free version can be understood as a special case of the noisy one with $\sigma_n = 0$.

<div style="display: inline-block;vertical-align:top; width: 48%;">

### Implementations in Python
```{python}
def sample_posterior(
    x, y, x_star, mean_func, cov_func, cov_args = {},
    sigma_n = 0.1, random_seed = -1, n_samples = 5
):
    k_xx = cov_func(get_r(x, x), **cov_args)
    k_xxs = cov_func(get_r(x, x_star), **cov_args)
    k_xsx = cov_func(get_r(x_star, x), **cov_args)
    k_xsxs = cov_func(get_r(x_star, x_star), **cov_args)
    
    I = np.identity(k_xx.shape[1])
    k_xx_noise = np.linalg.inv(k_xx + sigma ** 2 * I)
    kxsx_kxxNoise = np.matmul(k_xsx, k_xx_noise)
    # Eq.2.23, 24
    fsb = np.matmul(kxsx_kxxNoise, y_train_N)
    cov_fs = k_xsxs - np.matmul(kxsx_kxxNoise, k_xxs)
    random_seed = int(random_seed)
    if random_seed < 0:
        prng = np.random
    else:
        prng = np.random.RandomState(random_seed)
    out = prng.multivariate_normal(fsb, cov_fs, n_samples)
    return out, fsb, cov_fs
```

</div>
<div style="display: inline-block;vertical-align:top; width: 48%;">

### Implementations in R

```{r}
sample_posterior <- function(
  x, y, x_star, mean_func, cov_func, ...,
  sigma_n = 0.1, random_seed = -1, n_samples = 5
) {
  k_xx = cov_func(get_r(x, x), ...)
  k_xxs = cov_func(get_r(x, x_star), ...)
  k_xsx = cov_func(get_r(x_star, x), ...)
  k_xsxs = cov_func(get_r(x_star, x_star), ...)
  
  I = diag(1, dim(k_xx)[1])
  k_xx_noise = solve(k_xx + sigma_n ^ 2 * I)
  kxsx_kxxNoise = k_xsx %*% k_xx_noise
  # Eq.2.23, 24
  fsb = kxsx_kxxNoise %*% y
  cov_fs = k_xsxs - kxsx_kxxNoise %*% k_xxs
  random_seed <- as.integer(random_seed)
  if (random_seed > 0) set.seed(random_seed)
  return(MASS::mvrnorm(n_samples, fsb, cov_fs))
}
```

</div>

This time let's try to fit some points in R.
```{r, fig.width = 8, fig.height=8}
x_train <- c(-10, -8, -5, 3, 7, 15)
y_train <- c(-5, -2, 3, 3, 2, 5)
x_star <- seq(-20, 20, length = 200)

par(mfrow = c(3, 3))
l <- c(0.25, 1, 4)
v <- c(0.5, 2, 8)
for (i in seq(length(l))) {
  for (j in seq(length(v))) {
    dt <- sample_posterior(
      x_train, y_train, x_star, mean_func, matern_kernel, 
      l = l[i], v = v[j], n_samples = 10, random_seed = 100
    )
    matplot(x_star, t(dt), "l", col = rgb(0.3, 0.3, 0.3, 0.5), lty = 1,
            main = paste("l=", l[i], ",v=", v[j]), ylim = c(-15, 15))
    points(x = x_train, y = y_train, pch = 16, col = "red")
  }
}
```

Note, that when $\ell$ is small, it is easier for the predicted posterior to return to normal (prior), which is the mean function, 0 (see the points around x = 0). As $\ell$ increases, it becomes more and more likely the predicted $y_{x=0}$ to stay at the "local" value, which is provided by the nearest neighbor in `y`.

